---
title: "A2"
author: "Lindsay Filgas"
date: "10/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(mapview)
library(leaflet)
library(devtools)
library(censusapi)
```

```{r}
#2010
Sys.setenv(CENSUS_KEY="7305c43b014cedd32494ba8e15939faf9f133077")

smc_pop_2010 <-
  getCensus(
    name = "dec/pl",
    vintage = 2010,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "P001001"
  ) %>% 
  transmute(
    block =
      paste0(state,county,tract,block),
    pop = P001001
  )

dec_vars_2010 <-
  listCensusMetadata(
    name = "2010/dec/pl",
    type = "variables"
  )

smc_blocks_2010 <- blocks("CA", "San Mateo", year = 2010, progress_bar = F)

smc_pop_2010 <- smc_pop_2010 %>%
  mutate(GEOID10 = block) %>%
  select(-block)

smc_block_pop_2010 <-
  smc_pop_2010 %>%
  left_join(smc_blocks_2010 %>% select(GEOID10)) %>%
  st_as_sf()
```


```{r}
#2020
Sys.setenv(CENSUS_KEY="7305c43b014cedd32494ba8e15939faf9f133077")

smc_pop_2020 <-
  getCensus(
    name = "dec/pl",
    vintage = 2020,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "P1_001N"
  ) %>% 
  transmute(
    block =
      paste0(state,county,tract,block),
    pop = P1_001N
  )

dec_vars_2020 <-
  listCensusMetadata(
    name = "2020/dec/pl",
    type = "variables"
  )

smc_blocks_2020 <- blocks("CA", "San Mateo", year = 2020, progress_bar = F)

smc_pop_2020 <- smc_pop_2020 %>%
  mutate(GEOID20 = block) %>%
  select(-block)

smc_block_pop_2020 <-
  smc_pop_2020 %>%
  left_join(smc_blocks_2020 %>% select(GEOID20)) %>%
  st_as_sf()
```

2010 East Palo Alto Blocks
```{r}
epa_boundary <- places("CA", progress_bar = F) %>%
  filter(NAME == "East Palo Alto")

epa_blocks_2010 <-
  smc_block_pop_2010 %>%
  st_centroid() %>%
  .[epa_boundary, ] %>%
  #st_set_geometry(NULL) %>%
  st_drop_geometry() %>% 
  left_join(smc_block_pop_2010 %>% select(GEOID10)) %>%
  st_as_sf()
  #select(___) %>% 
  #left_join(___)

mapview(epa_boundary) + mapview(epa_blocks_2010, alpha.regions = 0, color = "red", lwd = 2)
```

2020 East Palo Alto Blocks
```{r}
epa_blocks_2020 <-
  smc_block_pop_2020 %>%
  st_centroid() %>%
  .[epa_boundary, ] %>%
  st_set_geometry(NULL) %>%
  left_join(smc_block_pop_2020 %>% select(GEOID20)) %>%
  st_as_sf()

mapview(epa_boundary) + mapview(epa_blocks_2020, alpha.regions = 0, color = "red", lwd = 2)
```

East Palo Alto Population Change 2010 to 2020

```{r}
epa_blocks_2020_intersect <- epa_blocks_2020 %>% 
  st_transform(26910) %>% 
  mutate(original_area = st_area(.)) %>%
  st_intersection(
    epa_blocks_2010 %>%
      select(GEOID10) %>% 
      st_transform(26910)
  ) %>%
  mutate(
    leftover_area = st_area(.),
    perc_area = as.numeric(leftover_area / original_area),
    pop = pop*perc_area
  )

epa_blocks_2020_reshaped <- epa_blocks_2020_intersect %>%
  st_drop_geometry() %>% 
  group_by(GEOID10) %>% 
  summarize(
    pop_2020 = sum(pop, na.rm=T) %>% round()
  )

projection <- epa_blocks_2020_reshaped

epa_pop_2010_2020 <- epa_blocks_2010 %>% 
  left_join(projection %>% select(GEOID10, pop_2020)) %>% 
  mutate(
    absolute_pop = (pop_2020 - pop) / as.numeric(st_area(.)) * 43560
  )

max <- max(abs(255))

absolute_pal <- colorNumeric(
  palette = "PiYG",
  domain = c(-255,255)
)

leaflet(epa_pop_2010_2020) %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  addPolygons(
    fillColor = ~absolute_pal(absolute_pop),
    label = ~round(absolute_pop, digits = 3),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.75,
    weight = 1
  #) %>% 
  #addLegend(
    #pal = ~absolute_pal(absolute_pop),
    #value = c(-255,255),
    #title = "Population Change"
  )
```

From this map, it is clear that there was the most population growth on the Palo Alto side of the Bayshore Highway, constituting the Western most part of East Palo Alto. There were lots of areas with minimal growth or slight decline in populations. This analysis makes many assumptions, including that population is evenly distributed across blocks, which was assumed when we calculated percent area and used this to estimate population. This data is also based solely on people who were able to fill out the census, which may not give an accurate representation of the populations of these areas.

