---
title: "A8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r}
library(tigris)
library(tidyverse)
library(tidycensus)
library(sf)
library(censusapi)
library(leaflet)
library(StatMatch)

Sys.setenv(CENSUS_KEY="7305c43b014cedd32494ba8e15939faf9f133077")

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()
```
Antioch BART station and PUMA of interest (01309)
```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_pumas,
    weight = 1,
    color = "gray",
    label = ~PUMACE10
  ) %>% 
  addMarkers(
    lng = -121.78033190708894,
    lat = 37.99565589827053
  ) %>% 
  addPolygons(
    data = bay_pumas %>% 
      filter(PUMACE10 == "01309")
  )
```

```{r}
pums_2014_2019 <- readRDS("pums_2014_2019_wts.rds")
  
pums_bart <- pums_2014_2019 %>%
  mutate(
    PWGTP = as.numeric(PWGTP),
    bart = ifelse(
      JWTR %in% c("4"),
      PWGTP,
      0
    )
  ) %>% 
  group_by(PUMA, year) %>% 
  summarize(
    pop = sum(PWGTP),
    bart = sum(bart)
  )
```

Map of Bay Area PUMAs
```{r}
pums_pal <- colorNumeric(
  palette = "YlOrRd",
  domain = pums_bart %>% 
    filter(year == 2018) %>% 
    pull(pop)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2018) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(pop),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": Population ", pop)
  )
```

Map of Bay Area BART riders
```{r}
pums_pal <- colorNumeric(
  palette = "GnBu",
  domain = pums_bart %>% 
    filter(year == 2018) %>% 
    pull(bart)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2018) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(bart),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": ", bart, " BART commute riders")
  )
```

```{r}
pums_bart_clean <-
  pums_bart %>% 
  select(-pop) %>% 
  pivot_wider(
    names_from = year,
    values_from = bart
  )
```

```{r}
obs_matrix <-
  pums_bart_clean %>% 
  ungroup() %>% 
  select(`2014`,`2015`,`2016`) %>% 
  as.matrix()

dist_matrix <- mahalanobis.dist(obs_matrix)

rownames(dist_matrix) <- pums_bart_clean$PUMA
colnames(dist_matrix) <- pums_bart_clean$PUMA

match <- dist_matrix["01309",] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    PUMA = rowname,
    match = "."
  ) %>% 
  right_join(
    pums_bart_clean
  ) %>% 
  arrange(match) %>% 
  .[1:11, ] %>% 
  left_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
  st_as_sf()
```

PUMA of interest (red) and 10 similar PUMAs (blue)
```{r}
leaflet() %>% 
  addTiles() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = match[1, ],
    color = "red",
    label = ~PUMA
  ) %>% 
  addPolygons(
    data = match[-1, ],
    label = ~PUMA
  )
```

```{r}
match_pumas <-
  match %>% 
  filter(!PUMA %in% c("01309")) %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>%
  group_by(
    year
  ) %>% 
  summarize(
    bart = mean(bart),
    PUMA = "Similar PUMAs"
  )

treatment_pumas <-
  match %>% 
  filter(PUMA %in% c("01309")) %>% 
  select(-match) %>% 
  st_drop_geometry() %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  )

rbind(
  treatment_pumas,
  match_pumas
) %>% 
  ggplot(
    aes(
      x = as.numeric(year),
      y = bart,
      color = PUMA
    )
  ) +
  geom_line() +
  geom_vline(xintercept = 2018, linetype = "dashed") +
  labs(
    title = "Antioch vs. control neighborhoods, BART ridership",
    x = "Year",
    y = "BART commute riders"
  )
```
From this graph it is clear that the PUMA of interest had a large increase in BART ridership in 2018 and onward compared to control PUMAs with similar trends prior to treatment. However, it is necessary to run a regression to understand if this is statistically significant. 
```{r}
transit_did <-
  match %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>% 
  mutate(
    year = year %>% as.numeric(),
    time = ifelse(year >= 2018, 1, 0),
    treated = ifelse(PUMA == "01309", 1, 0)
  )

did_reg <- lm(bart ~ treated*time, data = transit_did)

summary(did_reg)
```
From this, we can estimate that the Antioch BART station led to an estimated increase of 967 BART riders (the difference-in-difference effect). The positive effect size (in this case about 0.37) suggests that over time there is a large change in BART ridership. Our p-value is significantly low (well below 0.05), meaning the results here are statistically significant and we can confidently say that the addition of the Antioch BART station led to an increase in ridership in our PUMA of interest.

Of course, there are many assumptions made in this analysis. For one, the Antioch BART station was built in mid-2018, yet we are selecting 2018 as our year for treatment. The ridership data may have been collected before the BART sation was constructed and operational. Further, we only have one year of data after the treatment year to consider. Revisiting this question a few years down the line when there is more data to look at and we can get a sense of long term trends would be useful. The consistency in 2018 and 2019 ridership for our PUMA of interest is promising and may suggest that future years would have equally high ridership, though the number of people taking BART in nearby PUMAs may also increase.

Another assumption is that the PUMA we chose to evaluate is the best choice for understanding the impacts of this particular BART station. PUMAs are relatively large areas, which is a limitation to begin with. The Antioch BART station is not located within our PUMA, and the people most likely to use the BART station may live in a variety of PUMAs, meaning that its impact on ridership may be greater than what we are observing.

Additionally, we must think about the metric we are focusing on-- mode of transportation for commutes-- and whether this is the best way to measure the impact of a BART station. This does not capture use of BART for other kinds of travel. And there may be other important impacts to consider, such as the effect on nearby businesses and the local economy, movement of people into or away from the surrounding neighborhoods, effects on traffic and congestion, or effects on air pollution.

Choosing 10 similar PUMAs is also a choice we should consider the impacts of. From looking at our "match" dataframe, we can see that some PUMAs are a much better match than others, which could influence our results since we are averaging the control group. More importantly, many PUMAs have no data (or report 0 bar trips) for the years 2014-2018. We should interrogate whether this is an accurate measure and ridership significantly increased in these PUMAs in 2019, or whether there is something else going on here. It may be more useful to match these PUMAs based on other characteristics, such as the number of people who commute to work in general/via car, income, level of educational achievement, etc.

Despite all of this, our data suggests that BART ridership for commuting to work did increase with the addition of the Antioch BART station, which can help inform future infrastructure projects.