---
title: "Filgas_A4"
author: "Lindsay Filgas"
date: "10/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r}
library(tidyverse)
library(censusapi)
Sys.setenv(CENSUS_KEY = "7305c43b014cedd32494ba8e15939faf9f133077")

acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5",
    type = "variables"
  )

```


```{r}
census_race_categories <- 
  c(
    "White Alone",
    "Black or African American Alone",
    "American Indian and Alaska Native Alone",
    "Asian Alone",
    "Native Hawaiian and Other Pacific Islander Alone",
    "Some Other Race Alone",
    "Two or More Races"
  )

bay_educ_race <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "county:001",
      regionin = "state:06",
      vars = paste0("group(C15002",LETTERS[x],")")
    ) %>%
    select(ends_with("E") & -c(state,NAME)) %>%
    pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
    left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,NA,"education"),
        sep = "!!"
      ) %>% 
      filter(!is.na(education)) %>% 
      mutate(race = census_race_categories[x])
  })

```


```{r}
bay_race_total <-
  bay_educ_race %>% 
  group_by(race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(education = "Total")

bay_educ_race %>% 
  group_by(education, race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  rbind(bay_race_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = education %>% factor(levels = rev(c("Total",unique(bay_educ_race$education)))),
      y = estimate,
      fill = race %>% factor(levels = rev(unique(bay_educ_race$race)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Educational attainment",
    y = "Proportion of population 25 years or older",
    title = "Bay Area educational attainment by race",
    fill = "Race of individual"
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )  +
  guides(
    fill = guide_legend(
      reverse = T
    )
  )
```

From this graph, it is clear that a slightly disproportionate number of White individuals in the population attain higher levels of education, specifically bachelors degrees and higher. Perhaps the most striking result, though, is the disproportionately high number of individuals of "some other race alone" that do not achieve more than a high school education. They make up a relatively small portion of the total population but a very large portion of those with less than a high school diploma. The representation of individuals of some other race alone amongst the college graduates is quite small compared to their population size. It is important to note that this chart is only looking at individuals 25 years of age and older, so there may be recent educational attainment changes among those under 25 that are not yet reflected in this data.
